author: Holden Mui
desc: Holes in unit square, may the world burn
source: TSTST 2022/1
url: https://aops.com/community/p25516960
tags: [2023-08, find, bestpossible, zayin]

---

Let $n$ be a positive integer. Find the smallest positive integer $k$ such
that for any set $S$ of $n$ points in the interior of the unit square, there
exists a set of $k$ rectangles such that the following hold:
\begin{itemize}
  \item The sides of each rectangle are parallel to the sides of the unit square.
  \item Each point in $S$ is \emph{not} in the interior of any rectangle.
  \item Each point in the interior of the unit square but \emph{not} in $S$
    is in the interior of at least one of the $k$ rectangles.
\end{itemize}
(The interior of a polygon does not contain its boundary.)

---

We give the author's solution.
In terms of $n$, we wish find the smallest integer $k$ for which
$(0, 1)^2 \setminus S$ is always a union of $k$ open rectangles
for every set $S \subset (0, 1)^2$ of size $n$.

We claim the answer is $k = \boxed{2n+2}$.

The lower bound is given by picking
\[ S = \{(s_1, s_1), (s_2, s_2), \dots, (s_n, s_n)\}\]
for some real numbers $0 < s_1 < s_2 < \dots < s_n < 1$.
Consider the $4n$ points
\[S' = S + \{(\varepsilon, 0), (0, \varepsilon), (-\varepsilon, 0), (0, -\varepsilon)\} \subset (0, 1)^2\]
for some sufficiently small $\varepsilon > 0$.
The four rectangles covering each of
\[(s_1 - \varepsilon, s_1), (s_1, s_1 - \varepsilon), (s_n + \varepsilon, s_n), (s_n, s_n + \varepsilon)\]
cannot cover any other points in $S'$;
all other rectangles can only cover at most 2 points in $S'$,
giving a bound of
\[ k \geq 4 + \frac{\lvert S' \rvert - 4}{2} = 2n+2.\]

\begin{center}
\begin{asy}
draw(unitsquare);
size(8cm);
int n = 7;
real eps = 0.05;
for (int i = 1; i <= n; ++i) {
dot((i/(n+1), i/(n+1)));
dot((i/(n+1), i/(n+1)+eps), red);
dot((i/(n+1)+eps, i/(n+1)), red);
}
dot((1/(n+1), 1/(n+1)-eps), red);
dot((1/(n+1)-eps, 1/(n+1)), red);
\end{asy}
\end{center}

To prove that $2n+2$ rectangles are sufficient,
assume that the number of distinct $y$-coordinates
is at least the number of distinct $x$-coordinates.
Let
\[ 0 = x_0 < x_1 < \dots < x_m < x_{m+1} = 1, \]
where $x_1$, \dots, $x_m$ are the distinct $x$-coordinates of points in $S$,
and let $Y_i$ be the set of $y$-coordinates of points with $x$-coordinate $x_i$.
For each $1 \leq i \leq m$, include the $|Y_i| + 1$ rectangles
\[ (x_{i-1}, x_{i+1}) \times ((0, 1) \setminus Y_i) \]
in the union, and also include $(0, x_1) \times (0, 1)$ and $(x_m, 1) \times (0, 1)$;
this uses $m+n+2$ rectangles.
\begin{center}
\begin{asy}
size(8cm);
draw(scale(5)*unitsquare);
dot((1, 3));
dot((1, 1));
dot((2, 1));
dot((3, 2));
dot((3, 4));
dot((4, 1));

real eps = 0.1;
void rect(real l, real r, real b, real t, pen p) {
l += eps;
r -= eps;
b += eps;
t -= eps;
filldraw((l, b) -- (l, t) -- (r, t) -- (r, b) -- cycle, p + opacity(0.5));
}
void dottedrect(real l, real r, real b, real t) {
l += eps;
r -= eps;
b += eps;
t -= eps;
draw((l, b) -- (l, t) -- (r, t) -- (r, b) -- cycle, dashed);
}
rect(0, 2, 0, 1-eps, red);
rect(0, 2, 1, 3, red);
rect(0, 2, 3, 5, red);
rect(1, 3, eps, 1, green);
rect(1, 3, 1+eps, 5-eps, green);
rect(2, 4, 0, 2, orange);
rect(2, 4, 2, 4, orange);
rect(2, 4, 4, 5, orange);
rect(3, 5, 0+eps, 1, cyan);
rect(3, 5, 1, 5-eps, cyan);
dottedrect(eps, 1, eps, 5-eps);
dottedrect(4, 5-eps, 0, 5);
\end{asy}
\end{center}
All remaining uncovered points are between pairs of points
with the same $y$-coordinate and adjacent $x$-coordinates $\{x_i, x_{i+1}\}$.
There are at most $n-m$ such pairs by the initial assumption,
so covering the points between each pair with
\[ (x_i, x_{i+1}) \times (y - \varepsilon, y + \varepsilon)\]
for some sufficiently small $\varepsilon > 0$ gives a total of
\[(m+n+2) + (n-m) = 2n+2\]
rectangles.
