desc: Colombian points and lines
author: Ivan Guo (AUS)
source: IMO 2013/2
tags: [2018-08, combogeo, construct, free, wishful, dumb, design, troll, find, bestpossible, bet]
hardness: 20
url: https://aops.com/community/p5720110

---

A configuration of $4027$ points in the plane is called
\emph{Colombian} if it consists of $2013$ red points and $2014$ blue points,
and no three of the points of the configuration are collinear.
By drawing some lines, the plane is divided into several regions.
An arrangement of lines is \emph{good} for a Colombian configuration
if the following two conditions are satisfied:
\begin{enumerate}
  \ii[(i)] No line passes through any point of the configuration.
  \ii[(ii)] No region contains points of both colors.
\end{enumerate}
Find the least value of $k$ such that for any Colombian configuration
of $4027$ points, there is a good arrangement of $k$ lines.

---

The answer is $k \ge 2013$.

To see that $k = 2013$ is necessary,
consider a regular $4026$-gon and alternatively color the
points red and blue,
then place the last blue point anywhere
in general position (it doesn't matter).
Each side of the $4026$ is a red-blue line segment
which needs to be cut by one of the $k$ lines,
and each line can cut at most two of the segments.

Now, we prove that $k = 2013$ lines is sufficient.
Consider the convex hull of all the points.
\begin{itemize}
  \ii If the convex hull has any red points,
  cut that red point off from everyone else by a single line.
  Then, for each of the remaining $2012$ red points,
  break them into $1006$ pairs arbitrarily,
  and for each pair $\{A, B\}$ draw two lines
  parallel to $AB$ and close to them.

  \ii If the convex hull has two consecutive blue points,
  cut those two blue points off from everyone else by a single line.
  Then repeat the above construction
  for the remaining $2012$ blue points.
\end{itemize}
The end.
